How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-04-08 doc/paper.tex
worksmithing methods + results

We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. All occurrence information was downloaded for calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). 
This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya. This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction as discussed below.

All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence. After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the Earth) between any two occurrences of that species; this distance was measured in kilometers.

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

@@

We developed a discrete-time survival model to analyze our data in order to answer our question of how wee can we predict extinction risk at million year time scales. We considered four model variations: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but we include none of our historical covariates (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and we include our historical covariates (Model VP). See Table \ref{tab:model_def} for further explanation of how these models differ from each other. For a complete description of the statistical model used in this analysis, please see Section \ref{sec:model_desc}. For a description of how our models were implemented, please see Section \ref{sec:model_est}.

@@

We are interested in model performance into two contexts: in-sample and out-of-sample predictive performance i.e. forecasting.

@@

We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data). These in-sample measures, however, are not necessarily good estimates of the model's ability to predict data from the future. To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. Each fold represents a sequence of time points. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

Here is a description of how k-fold cross-validation for time series precedes \citep{Arlot2010,Bergmeir2016}. The model is fit first to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the first and second golds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. This process continues with the model fit to the first three folds used to forecast extinction probabilities for the fourth fold, and then the model fit to the first four folds being used to forecast the fifth fold.

Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}. This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance. AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of any two randomly selected species \citep{Fawcett2006,Mason2002}. AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, which means that values between 0.7 and 0.8 could then be considered ``fair.''

@@

See our code repository at https://github.com/psmits/trident for full code details. The entire analysis was coded in R and uses tidyverse and tidyverse adjacent tools such as \texttt{dplyr} \citep{dplyr}, and \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}. Additionally, all of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.

@@

The in-sample model comparisons are for determining their relative adequacy, or a model's ability to represent the data it was fit to. Comparison between the posterior predictive estimates of in-sample AUC for each of the four models demonstrates that, overall, all the models have approximately equal in-sample performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All the in-sample AUC estimates from our models are concentrated on an AUC value of 0.77. It is therefore hard to conclude that there is one ``best'' model which we can rely upon as they are all nearly functional equivalent.

@@

There are one or more time periods where posterior predictive in-sample AUC has a median value less than or equal to 0.5 for the various taxon/model combinations (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior predictive distribution of Foraminfera and Radiolaria for the VP model. Additionally, these periods of low model performance are rarer for the posterior predictive distribution of the VP model for calcareous nannoplankton and Dinoflagellates when compared to the other three models.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
  \caption{Comparison of in-sample forecasting performance measured by AUC for each of the four models, arranged over time and by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:auc_taxon_time}
\end{figure}

@@

Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}. The resulting distribution when all folds are combined is very multimodal by their very nature of being fit to and estimated from different data sets and amounts of data \citep{ESL}. We find the visual degree of this multimodality increases with model complexity (Fig. \ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time which allows for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

Comparison between the in-sample forecasting performance distributions of expected out-of-sample AUC reveals a similar range in plausible values for all models (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). Interestingly, the differences in the posterior predictive distributions of AUC for the models have decreased. For example, model VP not clearly better than either models V or CP (Fig. \ref{fig:fold_auc}), which were shown earlier to be obviously worse-performing models based on in-sample performance (Fig. \ref{fig:auc_hist}). These differences mean that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, the shapes of the posterior distributions means interpretations based only on the median values is incorrect -- the models are effectively indistinguishable in their out-of-sample forecast performance.

@@

We can also compare the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig. \ref{fig:fold_auc_taxon_time}). The time-series of posterior predictive values for each model are broadly congruent between each taxonomic group.

In the analysis of the posterior predictive distributions of AUC from the in-sample forecasts for the four models, we noted that there were time intervals where the models' predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior predictive distribution of AUC from the out-of-sample forecasts -- the major exception to this is Dinoflagellates, which for all four models has at least one time interval where the median the AUC of out-of-sample data were no better random. In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:fold_auc_taxon_time}
\end{figure}

We quantitatively compared our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}). This calculation means that differences close to 0 indicate complete congruence between in-sample and out-of-sample forecasts. A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, quantifying the broad congruence that the visual comparison of the two types of AUC time series suggest (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig. \ref{fig:oos_ins_diff}).

@@

The results of this paper set out our baseline ability to predict relative differences in extinction risk. We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations. Similarly, these models are expected to correctly forecast the rank order in extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on less data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.

>>>>

We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. Occurrence information was downloaded for calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). 
This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya. This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction as discussed below.

All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence as listed in the Neptune Database. After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the Earth) between any two occurrences of that species; this distance was measured in kilometers.

We also calculated the average global temperature of each 1 My bin as estimated from Magnesium/Calcium isotope ratios \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

@@

We used a discrete-time survival modelling framework in order to estimate how well we can predict extinction risk at one million year time scales. At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}. We considered four model variations of the predictors included in our model: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but none of our historical covariates are included (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and historical covariates are included (Model VP). See Table \ref{tab:model_def} for further explanation of how these models differ from each other. For a complete description of the statistical model used in this analysis, please see Section \ref{sec:model_desc}. For a full description of how our models were implemented, please see Section \ref{sec:model_est}.

@@

We are interested in our models' performance into two contexts: in-sample and out-of-sample predictive performance i.e. forecasting.

@@

We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data). In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}. To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. Each fold represents a sequence of time points. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

k-fold cross-validation for time series precedes as follows \citep{Arlot2010,Bergmeir2016}. First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. This process continues with the model fit to the first three folds combined is then used to forecast extinction probabilities for the fourth fold. Finally, the model is fit to the first four folds combined and then is used to forecast the fifth fold. The results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.

The relative adequacy of our four model variants were compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}. This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no difference in classification from random and 1 indicating perfect classification. AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of a randomly selected extinct-extant species pair \citep{Fawcett2006,Mason2002}. AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, so we interpret values between 0.7 and 0.8 could then be considered ``fair,'' and values between 0.6 and 0.7 as ``poor.''

@@

See our code repository at https://github.com/psmits/trident for full code details. This entire analysis was coded in R and uses tidyverse and tidyverse adjacent tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}. Additionally, all of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.

@@

The in-sample model comparisons are useful comparing the relative ability of our models to represent the data it was fit to. Comparison between the posterior distributions of in-sample AUC for each of the four models demonstrates that all of our models have approximately equal in-sample forecasting performance (Fig. \ref{fig:auc_hist}). The parameter rich model VP has the greatest median in-sample AUC when compared to the other three models, but there is substantial overlap in their posterior distributions. Additionally, while our parameter rich model VP is possibly the most adequately performing model, the difference or improvement to performance is minimal at best -- all four models have approximately equal in-sample AUC posterior distributions. All the in-sample AUC estimates from our models are concentrated on an AUC value of 0.77. It is therefore hard to conclude that there is one ``best'' model which we can rely upon as they are all nearly functional equivalent.

@@

There are one or more time periods where our posterior distribution of in-sample AUC has a median value less than or equal to 0.5 (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior estimates of in-sample AUC for Foraminfera and Radiolaria as fit by the VP model. In contrast, there are fewer periods of low model performance for calcareous nannoplankton and Dinoflagellates as estimated from our VP model than in those estimates from the other three model variations.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
  \caption{Comparison of in-sample forecasting performance measured by AUC for each of the four models, arranged over time and by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:auc_taxon_time}
\end{figure}

@@

Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}. The resulting distribution when all folds are combined is very multimodal by their very nature of being fit to and estimated from different data sets and amounts of data \citep{ESL}. We find that this multimodality increases with model complexity (Fig. \ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time which allows for a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

Comparison between the in-sample forecasting performance estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). Interestingly, the differences in posterior predictive distributions of AUC between the four model variations has decreased. For example, the VP model no longer has the greatest median AUC of the four models (Fig. \ref{fig:fold_auc}). These differences mean that the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, making interpretations bsaed only incorrect -- our estimates are the full posterior, not individual points. Because of this, the four models are effectively indistinguishable in their out-of-sample forecast performance (Fig. \ref{fig:fold_auc}).

@@

We can also compare the posterior predictive distribution of expected out-of-sample AUC over time and taxonomic group for each of the four models (Fig. \ref{fig:fold_auc_taxon_time}). In all cases, our four models are nearly identical in their out-of-sample forecasting performance over time and between each of the taxonomic groups.

In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts. The major exception to this pattern are our estimates for the Dinoflagellates, which has at least one time interval for all four models where the median AUC of the out-of-sample forecasts were no better random. In contrast, the only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with model V.
\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
  \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. See Table \ref{tab:model_def} for a description of each of the four models.}
  \label{fig:fold_auc_taxon_time}
\end{figure}

We compared the difference in our AUC estimates from the out-of-sample forecasts to the AUC estimates from our in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}). A difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts. A positive difference indicates that our out-of-sample forecasts are actually higher performing than our in-sample forecasts, while negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between our out-of-sample and in-sample forecasts are rare and tend to not form multimillion year patters, something that is consistent with the broad visual congruence between the in-sample and out-of-sample forecast performance (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). The only major multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model concentrated around 30 Mya (Fig. \ref{fig:oos_ins_diff}).

@@

The results of this paper set out our baseline ability to predict relative differences in extinction risk. We find that all of our model have an approximate 77\% to 79\% probability of correctly rank the extinction risk of two randomly selected in-sample observations. Similarly, these models are expected to correctly forecast the rank order in extinction risk of two randomly selected out-of-sample observations approximately 70\% to 80\% of the time. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic.
